hbm_paper_results

### 2024-03-06 eda/lifespan_eda.Rmd
getting numbers, figs

unique(cn.df.imp$site)
table(cn.df.imp$site, cn.df.imp$study)
sum(is.na(cn.df.imp$site))
max(cn.df.imp$age_days)/365.25
mean(cn.df.imp$age_days)/365.25
```

>>>>

# unique(cn.df.imp$site)
# table(cn.df.imp$site, cn.df.imp$study)
# sum(is.na(cn.df.imp$site))
```

Demographics

```{r}
#sample size, sex
length(unique(cn.df.imp$participant))
nrow(cn.df.imp)

table(cn.df.imp$sex)

#studies, sites
length(unique(cn.df.imp$study))
length(unique(cn.df.imp$site))

#age
min(cn.df.imp$age_days)/365.25
max(cn.df.imp$age_days)/365.25
mean(cn.df.imp$age_days)/365.25

abcd <- cn.df.imp %>%
  dplyr::filter(study=="ABCD")
length(unique(abcd$site))
```

### 2024-03-06 paper_prop_results.Rmd
getting numbers, figs

>>>>

Look at differences in M:F for subjects w very high or very low mean centiles
```{r}
feat_exts <- ratio.df %>%
  dplyr::select(-contains(c("diff", ".z"))) %>%
  mutate(across(pheno_list, ~ case_when(. < 0.25 ~ "low",
                                        . > 0.75 ~ "high", 
                                        TRUE ~ "mid")))

feat_exts_long <- feat_exts %>%
  pivot_longer(cols=c(pheno_list), names_to="pheno", values_to="centile_cat") %>%
   mutate(pheno_cat = as.factor(case_when(
    pheno %in% vol_list_global ~ "Global Volume",
    pheno %in% vol_list_regions ~ "Regional Volume",
    pheno %in% sa_list ~ "Regional SA",
    pheno %in% ct_list ~ "Regional CT",
    TRUE ~ NA_character_)))

feat_exts_long %>%
  ggplot() +
  geom_col(aes(x=centile_cat, y=pheno_cat, fill=sex, color=sex)) +
  facet_grid(dataset ~ prop)
```

### 2024-03-06 paper_replication_results.Rmd
getting numbers, figs

```{r}
data_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar/derivatives/ukb_perm"

@@

@@

ct.t.tests_to_plt.sum <- ct.t.tests_to_plt %>%
  group_by(hemi, region, label, bigger_group) %>%
  summarise(n_worse = sum(sig_fdr)) %>%
  pivot_wider(id_cols=c(hemi, label, region), names_from = bigger_group, values_from=n_worse) %>%
  mutate(cf.gamlss=ifelse(is.na(cf.gamlss), 0, cf.gamlss)) %>%
  mutate(cf.gamlss_minus_cf.gam_better=as.integer(cf.gam-cf.gamlss),
         percent_cf.gamlss_better=(cf.gam/10)*100) %>%
  ungroup()

@@

```{r}
dataset_names <- c(ukb_CN = "Raw Data", cf = "ComBat w/o Covar", cf.lm = "ComBat", cf.gam ="ComBat-GAM", cf.gamlss ="ComBatLS")
cent.diffs.v.plot <- perm_subj.df %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_cent_abs.diff, fill=perm), scale="count") +
  labs(y= "Mean Magnitude Centile Error", x = "ComBat Configurations") +
  scale_x_discrete(labels=c(dataset_names)) +
  guides(fill=guide_legend(title="Sampling \nPermutation")) + 
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat \nconfigurations replicated 10x")
cent.diffs.v.plot
# ggsave(filename="figs/centile_violin.png", cent.diffs.v.plot)
```

@@

#look by feature type
cent.subj.sex.test.df %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss" | dataset == "cf.lm") %>%
  mutate(dataset = factor(dataset, levels = c("cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  mutate(pheno_cat = factor(case_when(
    pheno %in% vol_list_global ~ "Global Volume",
    pheno %in% vol_list_regions ~ "Regional Volume",
    pheno %in% sa_list ~ "Regional SA",
    pheno %in% ct_list ~ "Regional CT",
    TRUE ~ NA_character_)),
    perm = as.factor(sub("perm-0", "", perm))) %>%
ggplot() +
  geom_histogram(aes(x=median_sex_diff, color=perm, fill=perm), alpha=0.5, position="identity") +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_grid(pheno_cat~dataset, scales="free", labeller = labeller(dataset = as_labeller(dataset_names))) +
  theme_linedraw(base_size=12) +
  labs(fill='Sampling \nPermutation', color='Sampling \nPermutation', x="Male - Female Median Centile Errors", y="Number of Features") 
```

@@

cent.sex.test.med %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_density(aes(x=median_of_median_sex_diff, fill=dataset, color=dataset), alpha=0.6, position = "identity") +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_wrap(~pheno_cat, scales="free")
```

plot subj. with averages in extremes - try w/ averages w/in pheno categories
```{r}
#recalc subject's true mean centiles
raw_mean_cent <- perm_subj.df %>%
  dplyr::filter(dataset=="cf") %>%
  mutate(mean_centile = mean_centile - mean_cent_diff,
         mean_z = mean_z - mean_z_diff) %>%
  dplyr::select(!c(Source_File, contains("diff"))) %>%
  mutate(dataset="raw")

@@

high_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=male_ratio)) +
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir="center") +
  geom_hline(yintercept= high_cent.sum[high_cent.sum$dataset == "raw" & high_cent.sum$perm=="perm-001",][["male_ratio"]], linetype="dashed")

@@

low_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=male_ratio)) +
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir="center")  +
  geom_hline(yintercept= low_cent.sum[low_cent.sum$dataset == "raw" & low_cent.sum$perm=="perm-001",][["male_ratio"]], linetype="dashed")

perm_subj.df %>%
  # dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_histogram(aes(x=mean_cent_diff, fill=sex, color=sex), position="identity", alpha=0.6, bins=60) +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_wrap(~dataset, scales="free")

# wilcox.test(mean_cent_diff ~ sex, data=perm_subj.df[dataset=="cf.gamlss",])
# wilcox.test(mean_cent_diff ~ sex, data=perm_subj.df[dataset=="cf.gam",])
# wilcox.test(mean_cent_diff ~ sex, data=perm_subj.df[dataset=="cf.lm",])

# perm_subj.df %>%
#   mutate(centile_raw = mean_centile - mean_cent_diff)
#   arrange(participant)
```

@@

high_test <- lapply(cf_test_list, one_samp_wilcox, df=high_cent.sum)
names(high_test) <- cf_test_list

low_test <- lapply(cf_test_list, one_samp_wilcox, df=low_cent.sum)
names(low_test) <- cf_test_list

high_test
low_test
```

>>>>

```{r}
data_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar/derivatives/ukb_perm"
base_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar"

@@

dataset_names <- c(ukb_CN = "Raw Data", cf = "ComBat w/o Covar", cf.lm = "ComBat", cf.gam ="ComBat-GAM", cf.gamlss ="ComBatLS")

@@

ct.t.tests_to_plt.sum <- ct.t.tests_to_plt %>%
  group_by(hemi, region, label, bigger_group) %>%
  summarise(n_worse = sum(sig_fdr)) %>%
  pivot_wider(id_cols=c(hemi, label, region), names_from = bigger_group, values_from=n_worse) %>%
  mutate(cf.gamlss=ifelse(is.na(cf.gamlss), 0, cf.gamlss)) %>%
  mutate(cf.gamlss_minus_cf.gam_better=as.integer(cf.gam-cf.gamlss),
         percent_cf.gamlss_better=(cf.gam/100)*100) %>%
  ungroup()

@@

```{r}
dataset_names <- c(ukb_CN = "Raw Data", cf = "ComBat w/o Covar", cf.lm = "ComBat", cf.gam ="ComBat-GAM", cf.gamlss ="ComBatLS")
cent.diffs.v.plot <- perm_subj.df %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_cent_abs.diff, fill=perm), scale="count") +
  labs(y= "Mean Magnitude Centile Error", x = "ComBat Configurations") +
  scale_x_discrete(labels=c(dataset_names)) +
  guides(fill=guide_legend(title="Sampling \nPermutation")) + 
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat \nconfigurations replicated 100x") +
  theme(legend.position="none")
cent.diffs.v.plot
# ggsave(filename="figs/centile_violin.png", cent.diffs.v.plot)
```

@@

#look by feature type
cent.subj.sex.test.df %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss" | dataset == "cf.lm") %>%
  mutate(dataset = factor(dataset, levels = c("cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  mutate(pheno_cat = factor(case_when(
    pheno %in% vol_list_global ~ "Global Volume",
    pheno %in% vol_list_regions ~ "Regional Volume",
    pheno %in% sa_list ~ "Regional SA",
    pheno %in% ct_list ~ "Regional CT",
    TRUE ~ NA_character_)),
    perm = as.factor(sub("perm-0", "", perm))) %>%
ggplot() +
  geom_histogram(aes(x=median_sex_diff, color=perm, fill=perm), alpha=0.3, position="identity") +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_grid(pheno_cat~dataset, scales="free", labeller = labeller(dataset = as_labeller(dataset_names))) +
  theme_linedraw(base_size=12) +
  theme(legend.position = "none") +
  labs(x="Male - Female Median Centile Errors", y="Number of Features") 
```

@@

cent.sex.test.med %>%
  # dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_density(aes(x=median_of_median_sex_diff, fill=dataset, color=dataset), alpha=0.6, position = "identity") +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_wrap(~pheno_cat, scales="free")

t.test.cent.sex.test.med <- cent.sex.test.med %>%
  group_by(dataset, pheno_cat) %>%
  summarise(t.test.p = tidy(t.test(median_of_median_sex_diff))$p.value) %>%
  mutate(p.val_fdr = p.adjust(t.test.p, method="fdr", n = nrow(cent.sex.test.med)),
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
t.test.cent.sex.test.med %>%
  ggplot() +
  geom_col(aes(x=pheno_cat, y=sig_fdr, fill=dataset), position="dodge")
```

plot subj. with averages in extremes - **try also using averages w/in pheno categories**
```{r}
#recalc subject's true mean centiles
raw_mean_cent <- perm_subj.df %>%
  dplyr::filter(dataset=="cf") %>%
  mutate(mean_centile = mean_centile - mean_cent_diff,
         mean_z = mean_z - mean_z_diff) %>%
  dplyr::select(!c(Source_File, contains("diff"))) %>%
  mutate(dataset="raw")

@@

high_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=male_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  geom_hline(yintercept= high_cent.sum[high_cent.sum$dataset == "raw" & high_cent.sum$perm=="perm-001",][["male_ratio"]], linetype="dashed")

@@

low_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  geom_hline(yintercept= low_cent.sum[low_cent.sum$dataset == "raw" & low_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed") +
  labs(title="subjects with low average centiles")

ext_cent.sum <- perm_subj.df2 %>%
  dplyr::filter(mean_centile < 0.25 | mean_centile > 0.75) %>%
  group_by(dataset, perm) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

ext_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=male_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  geom_hline(yintercept= ext_cent.sum[ext_cent.sum$dataset == "raw" & ext_cent.sum$perm=="perm-001",][["male_ratio"]], linetype="dashed")

perm_subj.df %>%
  # dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_histogram(aes(x=mean_cent_diff, fill=sex, color=sex), position="identity", alpha=0.6, bins=60) +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_wrap(~dataset, scales="free")
```

too many males in high centiles, too few males in low centiles (combat-GAM trends opposite to this pattern)

overall - too few men relative to women in extreme centiles

@@

high_test <- lapply(cf_test_list, one_samp_wilcox, df=high_cent.sum)
names(high_test) <- cf_test_list
high_test

low_test <- lapply(cf_test_list, one_samp_wilcox, df=low_cent.sum)
names(low_test) <- cf_test_list
low_test

ext_test <- lapply(cf_test_list, one_samp_wilcox, df=ext_cent.sum)
names(ext_test) <- cf_test_list
ext_test
```
